---
title: "DMTM Final Project - Data Analysis on Personal Spotify Data with R"
author: "Paul Dubois - pdubois6743@floridapoly.edu"
output: html_notebook
---

### Read in necessary packages

```{r Read in necessary packages}
library(readr)
library(DBI)
library(viridis)
library(lubridate)
library(tidyverse)
library(ggplot2)
library(spotifyr)
library(plotly)
library(knitr)
library("gghighlight")
library(jsonlite)
```


```{r}
PADstream00 <- fromJSON("StreamingHistory0.json", flatten = TRUE)
PADstream01 <- fromJSON("StreamingHistory1.json", flatten = TRUE)
PADstream02 <- fromJSON("StreamingHistory2.json", flatten = TRUE)

PAD_spotify <- rbind(PADstream00, PADstream01, PADstream02) #combining three JSON files to have only one dataset

PAD_spotify_2022 <- PAD_spotify %>% #formatting date using lubridate, and adding a date attribute as well as minutes and seconds
  as_tibble() %>% 
  mutate_at("endTime", ymd_hm) %>% 
  mutate(endTime = endTime - hours(6)) %>% 
  mutate(date = floor_date(endTime, "day") %>% as_date, seconds = msPlayed / 1000, minutes = seconds / 60) %>% 
  filter(date >= '2022-01-01' & date <= '2022-10-31')

str(PAD_spotify_2022)
```
> The total data is orginially from November 5, 2021 to November 5, 2022. However for this project, I will only focus on this year’s data, so I filter the original combined dataset into a new variable PAD_spotify_2022 which holds data from January 1, 2022 to October 31, 2022.

```{r}
#Calculating the total number of minutes of playback
totalMinutes <- sum(PAD_spotify_2022["msPlayed"])/1000/60
cat("You listened to a total of", totalMinutes, "minutes in 2022!")
```

> Calculating my total minutes listened was very simple.

```{r, warning = FALSE}
#Manually predicting 4 artists I think I've most listened to this year
artist_hours <- PAD_spotify_2022 %>% 
  group_by(artistName, date = floor_date(date, "month")) %>% 
  summarize(hours = sum(minutes)/60) 


ggplot(data = artist_hours) +
  geom_line(aes(x= date, y= hours, group = artistName, color = artistName)) +
  gghighlight(artistName == "Mac DeMarco" ||
              artistName == "Kanye West" ||
              artistName == "Bad Bunny" || 
              artistName == "Baby Keem") +
  scale_x_date(date_labels = "%b %Y", date_breaks = "2 month") +
  scale_color_manual(values = c("darkorange2", "blue", "red", "forest green")) +
  labs(title = "Playback hours per artist", subtitle = "Baby Keem vs. Kanye West") +
  xlab("") +
  theme_light()
```
> For most of the ggplots, new transformations were needed to be made to the original `PAD_spotify_2022`. In this ggplot I am plotting all artists listened to, then visualizing and highlighting those with the highest spikes in hours listened.

```{r, warning = FALSE}
#top 5 most listened to artists
PAD_most_listened <- PAD_spotify_2022 %>% 
  group_by(artistName) %>% 
  mutate(artistName = iconv(artistName, to = "UTF-8")) %>% 
  summarize(listenedHours = sum(minutes)/60) %>% 
  arrange(desc(listenedHours)) %>% 
  slice(1:5)

PAD_most_listened
```

> A tibble representation of my top 5 artists categorized by the total number of hours I listened to them.

```{r, warning = FALSE}
ggplot(data = PAD_most_listened) +
  geom_segment(aes(x= artistName, xend= artistName, y= 0, yend = listenedHours)) +
  scale_color_viridis(discrete = TRUE, guide = FALSE, option="C") +
  theme_light(base_size=12, base_family="HiraKakuProN-W3") +
  theme (
    panel.grid.major.x = element_blank(),
    panel.border = element_blank(),
  geom_point(mapping=aes(x = artistName, y = listenedHours,color = artistName, color = "cyan3")) +
    axis.ticks.x = element_blank(),
  ) +
  labs(title = "Top 5 most listened artists") +
  xlab("") +
  ylab("Total Hours Listened") +
  coord_flip()
```
> A visualization representation of my top 5 artists categorized by the total number of hours I listened to them.

```{r, warning = FALSE}
#top 5 most listented to songs
PAD_top_songs <- PAD_spotify_2022 %>% 
  group_by(trackName) %>% 
  summarize(minutes= sum(minutes)) %>% 
  arrange(desc(minutes)) %>% 
  slice(1:5)

PAD_top_songs
```

> A tibble representation of my top 5 artists categorized by the total number of hours I listened to them.

```{r, warning = FALSE}
ggplot(data = PAD_top_songs) +
  geom_segment(aes(x= trackName, xend= trackName, y= 0, yend= minutes)) +
  geom_point(mapping=aes(x= trackName, y= minutes, color= trackName)) +
  scale_color_viridis(discrete = TRUE, guide = FALSE, option="D") +
  theme_light(base_size=12, base_family="HiraKakuProN-W3") +
  theme (
    panel.grid.major.x = element_blank(),
    panel.border = element_blank(),
    axis.ticks.x = element_blank(),
  ) +
  labs(title = "Top 5 listened songs") +
  xlab("") +
  ylab("Total Minutes Listened") +
  coord_flip()

```
> A visualized representation of my top 5 songs categorized by the total number of hours I listened to them.

```{r}
PAD_stream_in_hours <- PAD_spotify_2022 %>% 
  group_by(date = floor_date(date, "week")) %>% 
  summarize(hours = sum(minutes)/60) %>% 
  arrange(date)

PAD_stream_in_hours
```
> A tibble representation of my time listened categorized by the week of the year.

```{r}
ggplot(data = PAD_stream_in_hours) +
  geom_col(mapping=aes(x = date, y = hours)) + xlab("Week") + ylab("Hours") +
  labs(title="Playback Hours each Week", subtitle="Breaking down Total Minutes Listened into Hours per Week") 
```

> A visualized representation of my time listened categorized by the week of the year.

```{r}
PAD_dayHour <- PAD_spotify_2022 %>% 
  group_by(date, hour = hour(endTime), weekday = wday(date, label = TRUE)) %>% 
  summarize(hoursListened = sum(minutes)/60)

PAD_weekHour <- PAD_dayHour %>% 
  group_by(weekday, hour) %>% 
  summarize(hours = sum(hoursListened))

PAD_dayHour
PAD_weekHour
```

```{r}
ggplot(data = PAD_weekHour) +
  geom_line(mapping=aes(x= hour, y= hours, color= weekday)) +
  scale_color_viridis(discrete= TRUE, option= "H") +
  scale_x_continuous(breaks= seq(0, 24, 2)) +
  scale_y_continuous(breaks= seq(0, 15, 2)) +
  scale_fill_gradient(low= "darkolivegreen1", high= "cyan4") +
  labs(title= "Listening Behaviors", subtitle= "Paul's Weekly Activity",
       x= "Time of Day", y= "Frequency measured by Total Hours") +
  theme_light() 

```

```{r }
PAD_dayHour %>% 
  group_by(weekday, hour) %>% 
  summarize(hours = sum(hoursListened)) %>% 
  ggplot(aes(x= hour, y= weekday, fill= hours)) +
  geom_tile() +
  scale_x_continuous(breaks= seq(0, 24, 2)) +
  scale_fill_gradient(low= "skyblue", high= "darkblue") +
  labs(title= "Spotify time during the week", subtitle= "Activity from 0 to 24 hours", x= "Time of Day", y= "") +
  theme_light()
```
```


