---
title: "FARS Project"
author: "Paul Dubois"
output: html_notebook
---


```{r}
library(tidyverse)
library(janitor)
library(scales)
```

# Function to Clean Column Names of `.xls` Files Gathered From NHTSA's FARS Online Query

```{r}
source("clean_csv.R")
```


## Yearly Total Crashes 

```{r}
df_crash_year <- clean_csv("FatalAccidents/Crash_per_Month_2010-2020.csv")
```

## Wrangling

```{r}
tidy_year <- df_crash_month %>% 
  pivot_longer(cols = 3:length(df_crash_month)-1, names_to = "Month", values_to = "Fatal_Crashes") %>% 
  clean_names()
```

* Check data

```{r}
tidy_year
```

### Making a line plot to show the trend

```{r}
tidy_year %>% 
  group_by(year) %>% 
  summarise(avg_crashes = mean(fatal_crashes)) %>% 
  ggplot(aes(x = year, y = avg_crashes)) +
  geom_line(color = "red") +
  labs(title = "How has the Average Number of Crashes Changed in the Last Decade?",
       subtitle = "2010 - 2020") +
  xlab("") + ylab("Average Crashes") +
  scale_x_discrete(limits = c(2010:2020))
```


## Car Manufacturer Wrangling & Visualization

```{r}
df_makes <- clean_csv("FatalAccidents/Vehicle_Make_2010-2020.csv")
```

### Tidy the data

```{r}
#Tidy Data
df_makes %>% 
  pivot_longer(cols = "ASTON_MARTIN":"Unknown_Make", names_to = "Make", values_to = "Fatal_Crashes")
```

```{r}
df_makes %>% 
  pivot_longer(cols = "ASTON_MARTIN":"Unknown_Make", names_to = "Make", values_to = "Fatal_Crashes") %>% 
  group_by(Make) %>% 
  summarise(total_crashes = sum(Fatal_Crashes, na.rm = T)) %>% 
  arrange(desc(total_crashes)) %>% 
  head(10) %>% 
  mutate(Make = str_replace_all(Make, "\\_|\\-", " ")) %>%  # This line RENAMES to look pretty for visualizations
  ggplot() +
    geom_col(aes(x = fct_reorder(Make,total_crashes), y = total_crashes), fill = "darkblue") +
    theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1, face = "bold")) +
    xlab(label = "Car Manufacturer") + ylim(0, 90000) +
    labs(title = "Which Manufacturer Has The Most Fatal Crashes from 2010 to 2020?", subtitle = "Sum of Total Crashes from 2010 to 2020.")
```

```{r}
df_makes %>% 
  group_by(Year) %>% 
  summarise(total_crashes = sum(Total))
```

```{r}
total_crashes_decade <- df_makes %>% 
  group_by(Year) %>% 
  summarise(total_crashes = sum(Total)) %>% 
  summarise(decade_crashes = mean(total_crashes)) %>% 
  pull
cat("Average crashes from 2010 - 2020:", total_crashes_decade)
```

```{r}
df_makes %>% 
  pivot_longer(cols = "ASTON_MARTIN":"Unknown_Make", names_to = "Make", values_to = "Fatal_Crashes") %>% 
  group_by(Make) %>% 
  summarise(total_crashes = sum(Fatal_Crashes, na.rm = T)) %>% 
  arrange(desc(total_crashes)) %>% 
  head(10) %>% 
  ggplot() +
    geom_col(aes(x = fct_reorder(Make,total_crashes), y = total_crashes), fill = "darkblue") +
    geom_hline(yintercept = total_crashes_decade) +
    theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1, face = "bold")) +
    xlab(label = "") + scale_y_continuous(name="Crashes", labels = comma) +
    labs(title = "Total Fatal Accidents Compared to the Top 10", subtitle = "Compared to the Average Number of Fatal Road Crashes")
```

## Experimenting with Word Cloud 

```{r}
library(wordcloud2)
tidy_manufacturer <- df_makes %>% 
  pivot_longer(cols = "ASTON_MARTIN":"Unknown_Make", names_to = "Make", values_to = "Fatal_Crashes") %>% 
  group_by(Make) %>% 
  summarise(total_crashes = sum(Fatal_Crashes, na.rm = T)) %>% 
  arrange(desc(total_crashes)) %>% 
  mutate(Make = str_replace(Make, pattern = "\\_", " "))
```

```{r}
wordcloud2(tidy_manufacturer, size = 0.6, gridSize = 4, color = "random-dark")
```


```{r}
df_makes %>% 
  clean_names() %>% 
  pivot_longer(cols = 3:length(df_makes)-1, names_to = "make", values_to = "fatal_crashes") %>% 
  group_by(make) %>% 
  summarise(total_crashes = sum(fatal_crashes, na.rm = T)) %>% 
  arrange(desc(total_crashes)) %>% 
  mutate(make = str_replace(make, pattern = "\\_", " ")) %>% 
  wordcloud2(size = 1)
```


## First Harmful Event Wrangling & Visualization

```{r}
df_event <- clean_csv("FatalAccidents/First_Harmful_Event_2010-2020.csv")
```

### Tidy the data

```{r}
event_summary <- df_event %>% 
  pivot_longer(cols = 3:length(df_event)-1, names_to = "Event", values_to = "Fatal_Crashes") %>% 
  group_by(Event) %>% 
  summarise(total_crashes = sum(Fatal_Crashes)) %>% 
  arrange(desc(total_crashes)) %>% 
  mutate(normalized_total = (total_crashes)/1000)
event_summary
```

### Tidy the data

```{r} 
event_names <- df_event %>% 
  pivot_longer(cols = 3:length(df_event)-1, names_to = "Event", values_to = "Fatal_Crashes") %>% 
  mutate(Event = str_replace_all(Event, "\\_|\\-", " "))
```

### Experiment with Treemaps


```{r}
library(treemapify)
event_names %>% 
  group_by(Event) %>% 
  summarise(total_crashes = sum(Fatal_Crashes)) %>% 
  arrange(desc(total_crashes)) %>% 
  mutate(normalized_total = (total_crashes)/1000) %>% 
  head(12) %>% 
  ggplot(aes(fill = Event, area = normalized_total, label = Event)) +
    geom_treemap() +
    geom_treemap_text(colour = "white", place = "centre") +
    labs(title = "First Harmful Event Visual") +
    theme(legend.position = "none")
```

### Experiment with Pie Charts

```{r}
event_names %>% 
  group_by(Event) %>% 
  summarise(total_crashes = sum(Fatal_Crashes)) %>% 
  arrange(desc(total_crashes)) %>% 
  mutate(normalized_total = (total_crashes)/1000) %>% 
  head(10) %>%
  ggplot(aes(x = normalized_total, y = Event, 
             fill = Event)) +
    geom_bar(width = 1, 
             stat = "identity", 
             color = "black") +
    coord_polar("y", 
                start = 0, 
                direction = -1) +
    theme_void()
```

## Atmospheric Conditions Wrangling & Visualization

```{r}
df_atmospheric_cond <- clean_csv("FatalAccidents/Atmospheric_Conditions_2010-2020.csv")
```

### Tidy data

```{r}
tidy_atmospheric_df <- df_atmospheric_cond %>% 
  pivot_longer(cols = 3:length(df_atmospheric_cond)-1, names_to = "Atmosphere", values_to = "Fatal_Crashes") %>% 
  mutate(Atmosphere = str_replace_all(Atmosphere, "\\_|\\-", " ")) %>% 
  clean_names()
sample_n(tidy_atmospheric_df,15)
```


### Experiment with Treemapify

```{r}
library(treemapify)
tidy_atmospheric_df %>% 
  group_by(atmosphere) %>% 
  summarise(total_crashes_atmo = sum(fatal_crashes)) %>% 
  arrange(desc(total_crashes_atmo)) %>% 
  ggplot(aes(fill = atmosphere, area = total_crashes_atmo, label = atmosphere)) +
    geom_treemap() +
    geom_treemap_text(colour = "white", place = "centre") +
    labs(title = "Atmospheric Conditions") +
    theme(legend.position = "none")
```



```{r}
thresh_freq <- 1020.403
small_class <- tidy_body_class %>%
  group_by(body_class) %>%
  summarise(avg_crashes = mean(fatal_crashes), 
            category = ifelse(avg_crashes > thresh_freq, body_class, "Other")) %>%
  group_by(category) %>%
  summarise(avg_crashes = mean(avg_crashes))
```


### Lollipop Plot


```{r}
decade_average <- 1020.403
class_graph <-  ggplot(data = small_class) +
  geom_segment(aes(x = fct_reorder(category, avg_crashes), xend = fct_reorder(category, avg_crashes) ,y = 0, yend = avg_crashes), color = "gray", size = 1) +
  geom_point(aes(x = fct_reorder(category, avg_crashes), y = avg_crashes), color = "black", size = 3) +
  geom_hline(yintercept = decade_average, color = "blue") +
  labs(title = "What Are The Most Dangerous Body Types?") +
  xlab(label = "") + ylab(label = "Average Crashes from 2010 to 2020") +
  theme_classic() +
  coord_flip()

class_graph

```
## Nationwide Crashes Wrangling & Visualization

```{r}
df_state_crashes <- clean_csv("FatalAccidents/Crashes_State.csv")
```
### Read data

```{r}
clean_state_df <- df_state_crashes %>% 
  pivot_longer(cols = 3:length(df_state_crashes)-1, names_to = "State", values_to = "crashes") %>% 
  clean_names() %>% 
  mutate(state = str_replace_all(state, "\\_|\\-", " ")) %>% 
  select(year, state, crashes, total)
```


### Unfortunate Obstacle

* The package `usmap` only plots the 50 states, disregarding _District of Columbia_
* To plot the data, `usmap` requires to have the `state_abbr`, `state_name`, and `value` in tidy format

```{r}
library(usmap)
no_dc <- clean_state_df %>% 
  filter(!state == "District of Columbia") 
```

* Create a list of state abbreviations
* Replicate down all observations

```{r}
state_names <- state.abb[which(state.name == no_dc$state)]
state_names <- state_names[1:50]
no_dc$state_abbr <- rep(state_names, 11)
no_dc 
```

* Check that all 50 unique state abbreviations and state names are still present

```{r}
unique(no_dc$state)
unique(no_dc$state_abbr)
```

* Visualize distribution of crashes for each year

```{r}
ggplot(no_dc) +
  geom_histogram(aes(x = crashes), fill = "skyblue") +
  facet_wrap(~year) +
  labs(title = "Distribution of Crashes for Each Year", subtitle = "Decade of 2010 to 2020")
```


### Wrangle For Visualization

```{r}
wrangled_states <- no_dc %>% 
  group_by(state, state_abbr) %>% 
  summarise(decade_crashes = sum(crashes))
```

## (Very) Redundant Process to Make Gif of US Map and Crashes Per State

* Create a subset dataset for each plot
* I attempted to automate with `for` loops and functions but failed

```{r}
states_2010 <- no_dc %>% 
  mutate(year = as.numeric(year, na.rm = T),
         crashes = log(crashes)) %>% 
  filter(year == 2010)

states_2011 <- no_dc %>% 
  mutate(year = as.numeric(year, na.rm = T),
         crashes = log(crashes)) %>% 
  filter(year == 2011)

states_2012 <- no_dc %>% 
  mutate(year = as.numeric(year, na.rm = T),
         crashes = log(crashes)) %>% 
  filter(year == 2012)

states_2013 <- no_dc %>% 
  mutate(year = as.numeric(year, na.rm = T),
         crashes = log(crashes)) %>% 
  filter(year == 2013)

states_2014 <- no_dc %>% 
  mutate(year = as.numeric(year, na.rm = T),
         crashes = log(crashes)) %>% 
  filter(year == 2014)

states_2015 <- no_dc %>% 
  mutate(year = as.numeric(year, na.rm = T),
         crashes = log(crashes)) %>% 
  filter(year == 2015)

states_2016 <- no_dc %>% 
  mutate(year = as.numeric(year, na.rm = T),
         crashes = log(crashes)) %>% 
  filter(year == 2016)

states_2017 <- no_dc %>% 
  mutate(year = as.numeric(year, na.rm = T),
         crashes = log(crashes)) %>% 
  filter(year == 2017)

states_2018 <- no_dc %>% 
  mutate(year = as.numeric(year, na.rm = T),
         crashes = log(crashes)) %>% 
  filter(year == 2018)

states_2019 <- no_dc %>% 
  mutate(year = as.numeric(year, na.rm = T),
         crashes = log(crashes)) %>% 
  filter(year == 2019)

states_2020 <- no_dc %>% 
  mutate(year = as.numeric(year, na.rm = T),
         crashes = log(crashes)) %>% 
  filter(year == 2020)
```

* Assign each plot to a variable

```{r}
p10 <- plot_usmap(data = states_2010, values = "crashes") + 
  scale_fill_continuous(low = "beige", high = "forestgreen", name = "Total Crashes", label = scales::comma) + 
  labs(title = "Fatal Crashes for 2010", subtitle = "") +
  theme(legend.position = "right")

p11 <- plot_usmap(data = states_2011, values = "crashes") + 
  scale_fill_continuous(low = "beige", high = "forestgreen", name = "Total Crashes", label = scales::comma) + 
  labs(title = "Fatal Crashes for 2011", subtitle = "") +
  theme(legend.position = "right")

p12 <- plot_usmap(data = states_2012, values = "crashes") + 
  scale_fill_continuous(low = "beige", high = "forestgreen", name = "Total Crashes", label = scales::comma) + 
  labs(title = "Fatal Crashes for 2012", subtitle = "") +
  theme(legend.position = "right")

p13 <- plot_usmap(data = states_2013, values = "crashes") + 
  scale_fill_continuous(low = "beige", high = "forestgreen", name = "Total Crashes", label = scales::comma) + 
  labs(title = "Fatal Crashes for 2013", subtitle = "") +
  theme(legend.position = "right")

p14 <- plot_usmap(data = states_2014, values = "crashes") + 
  scale_fill_continuous(low = "beige", high = "forestgreen", name = "Total Crashes", label = scales::comma) + 
  labs(title = "Fatal Crashes for 2014", subtitle = "") +
  theme(legend.position = "right")

p15 <- plot_usmap(data = states_2015, values = "crashes") + 
  scale_fill_continuous(low = "beige", high = "forestgreen", name = "Total Crashes", label = scales::comma) + 
  labs(title = "Fatal Crashes for 2015", subtitle = "") +
  theme(legend.position = "right")

p16 <- plot_usmap(data = states_2016, values = "crashes") + 
  scale_fill_continuous(low = "beige", high = "forestgreen", name = "Total Crashes", label = scales::comma) + 
  labs(title = "Fatal Crashes for 2016", subtitle = "") +
  theme(legend.position = "right")

p17 <- plot_usmap(data = states_2017, values = "crashes") + 
  scale_fill_continuous(low = "beige", high = "forestgreen", name = "Total Crashes", label = scales::comma) + 
  labs(title = "Fatal Crashes for 2017", subtitle = "") +
  theme(legend.position = "right")

p18 <- plot_usmap(data = states_2018, values = "crashes") + 
  scale_fill_continuous(low = "beige", high = "forestgreen", name = "Total Crashes", label = scales::comma) + 
  labs(title = "Fatal Crashes for 2018", subtitle = "") +
  theme(legend.position = "right")

p19 <- plot_usmap(data = states_2019, values = "crashes") + 
  scale_fill_continuous(low = "beige", high = "forestgreen", name = "Total Crashes", label = scales::comma) + 
  labs(title = "Fatal Crashes for 2019", subtitle = "") +
  theme(legend.position = "right")

p20 <- plot_usmap(data = states_2020, values = "crashes") + 
  scale_fill_continuous(low = "beige", high = "forestgreen", name = "Total Crashes", label = scales::comma) + 
  labs(title = "Fatal Crashes for 2020", subtitle = "") +
  theme(legend.position = "right")
```

* Create disk that will be used to save the plots

```{r}
dir_out <- file.path(tempdir(), "nation_crashes")
dir.create(dir_out, recursive = TRUE)
```

* Then plots will be written to disk as static `.png` files using this function.

```{r}
save_gg <- function(graph) {
  fp <- file.path(dir_out, paste0("year", ".png"))
  
  ggsave(plot = graph, 
         filename = fp, 
         device = "png")
}
```

* With the plots written to disk, use magick to read in the image data, file will appear in working directory as png.
* Took the 11 plots and used an onnline gif maker to create an appealing gif

```{r}
library(magick)
save_gg(p20)
imgs <- list.files(dir_out, full.names = TRUE)
img_list <- lapply(imgs, image_read)

## join the images together
img_joined <- image_join(img_list)

## animate at 2 frames per second
img_animated <- image_animate(img_joined, fps = 2)

## view animated image
img_animated

## save to disk
image_write(image = img_animated,
            path = "nation_crashes.png")
```

## Interactive US Map Plot With Total Crashes from 2010 to 2020

```{r}
library(usmap)
library(plotly)
g <- plot_usmap(data = wrangled_states, values = "decade_crashes") + 
  scale_fill_continuous(low = "skyblue", high = "darkblue", name = "Total Crashes", label = scales::comma) + 
  labs(title = "Fatal Crashes from 2010 to 2020", subtitle = "Aggregated data from 2010 to 2020") +
  theme(legend.position = "right")

g
```


## Florida County Wrangling & Visualization

```{r}
df_fl_counties <- clean_csv("FatalAccidents/Florida_county_crashes.csv")
```

## Tidy data

* Notice that the dataset we queried from NHTSA decided to keep _Florida_ before each name of counties
* Let's separate to keep the state name (in case we need it) and county names!

```{r}
tidy_fl_counties <- df_fl_counties %>% 
  pivot_longer(cols = 3:length(df_fl_counties)-1, names_to = "County", values_to = "Crashes") %>% 
  mutate(Year = as.numeric(Year, na.rm = T)) %>%
  mutate(County = str_remove_all(County, pattern = "Florida\\_|\\_|-")) %>% 
  select(Year, County, Crashes, Total) %>% 
  clean_names()

tidy_fl_counties
```


```{r}
county_decade <- tidy_fl_counties %>% 
  group_by(county) %>% 
  summarise(decade_crashes = sum(crashes)) 

county_decade$state_abbr <- rep("FL")
county_decade$state <- rep("Florida")
```

## Florida County Map

```{r}
county_decade$county[50] <- "Palm Beach"
source("county_fips.R")
```


```{r}
county_decade$fips <- fips_code
```

```{r}
county_decade
```


```{r}
plot_usmap("counties", include = c(.south_atlantic), exclude = c("13","45","37","51","54","10", "24", "11"),
           data = county_decade, values = "decade_crashes") +
  scale_fill_continuous(low = "skyblue", high = "darkblue", name = "Total Crashes", label = scales::comma) + 
  labs(title = "Which County has the Most Dangerous Drivers?", subtitle = "Aggregated data from 2010 to 2020") +
  theme(legend.position = "right")
```

> Why isn't Miami-Dade County being shown?

## Car Body-Class Wrangling & Visualization

```{r}
df_body_class <- clean_csv("FatalAccidents/Crash_Body_Class_2010-2020.csv")
```

### Tidy data

```{r}
tidy_body_class <- df_body_class %>% 
  pivot_longer(cols = 3:length(df_body_class)-1, names_to = "Body_Class", values_to = "Fatal_Crashes") %>% 
  mutate(Body_Class = str_replace_all(Body_Class, "\\_|\\-", " ")) %>% 
  select(Year, Body_Class, Fatal_Crashes, Total) %>% 
  clean_names() 
```

### Summary Statistics

```{r}
options(scipen = 999)
avg_crashes <- tidy_body_class %>% 
  mutate(
    fatal_crashes = as.numeric(fatal_crashes)
  ) %>% 
  group_by(body_class) %>% 
  summarise(decade_avg = mean(fatal_crashes))
```

* Find the average amount of fatal crashes each year

```{r}
mean(avg_crashes$decade_avg)
```

### Plot Cleaning

* The following summaries creates a wrangle of the `tidy_body_class` to combine the minimal counts of crashes to an `other` category
* This gives a much cleaner look to the graph

```{r}
thresh_freq <- 1020.403
small_class <- tidy_body_class %>%
  group_by(body_class) %>%
  summarise(avg_crashes = mean(fatal_crashes), 
            category = ifelse(avg_crashes > thresh_freq, body_class, "Other")) %>%
  group_by(category) %>%
  summarise(avg_crashes = mean(avg_crashes))
```


### Lollipop Plot

* I have really enjoyed creating these plots, they look good and are compelling

```{r}
decade_average <- 1020.403
class_graph <-  ggplot(data = small_class) +
  geom_segment(aes(x = fct_reorder(category, avg_crashes), xend = fct_reorder(category, avg_crashes) ,y = 0, yend = avg_crashes), color = "gray", size = 1) +
  geom_col(aes(x = fct_reorder(category, avg_crashes), y = avg_crashes), fill = "lightblue") +
  geom_hline(yintercept = decade_average, color = "blue") +
  labs(title = "What Are The Most Dangerous Body Types?") +
  xlab(label = "") + ylab(label = "Average Crashes from 2010 to 2020") +
  theme_classic() +
  coord_flip()

class_graph
```


## Conclusions

* This project was a sole attempt to show what I have learned this semester
* Given the time we have spent, reading, cleaning, wrangling and visualizing data in class; I have taken a noticeable step forward in the world of data analysis
* There is a grand diversity of plots, column, histogram, 
